A global sensitivity analysis of model uncertainty in aeroelastic wind turbine models

### 2019-12-17 Papers/Torque_Paper/article.tex
updated article text

For this case of a rigid turbine, with a uniform inflow field, the main model uncertainties that arise in the BEM formulation that we will discuss in this work are the following \cite{Hansen1993}:
\begin{itemize}
\item The use of (2D) \textit{airfoil polars}; uncertainty arises not only because the actual flow along the blade is 3D, but also because the 2D polars itself can be inaccurate, either when obtained from measurement or from 2D codes like Xfoil;
%\item Sheared inflow
\item The assumption of locally 2D flow along the blade is also invalid at the tip; this is typically corrected via \textit{a tip loss model};
\item Three-dimensional flow effects also cause stall delay at the root; this is typically modelled via \textit{a dynamic stall model }and time constant;
\item Yawed inflow is not naturally included in BEM and typically incorporated via \textit{a wake correction}.

@@

\subsection{Geometric uncertainty}
We use Non-Uniform Rational Basis Splines (NURBS) \cite{rogers2000} to perturb the geometrical parameters of the turbine blade, such as the reference chord and twist curves. The main advantage of using NURBS is that it provides great flexibility to approximate a large variety of curves with a limited number of control points. Further, the set of control points and knots can be directly manipulated to control the smoothness and curvature.

Use of NURBS in wind-turbine blades (airfoils) optimization: \cite{Ribeiro2012,Bottasso2014}.

\subsubsection{NURBS based perturbation}
The value of NURBS curve at each location $x$ is computed using a weighted sum of $N$ basis functions (or B-splines):
\begin{equation}\label{NURB_curve}
S(x)  = \sum_{i=1}^{N} c_i B_{i,p}(x),
\end{equation}
where $S(x)$ is the value of the curve at location $x, c_i$ is the weight of control point $i$ and $B_{i,p}(x)$ is the value of B-spline corresponding to the $i$-th control point at $x$. The subscript $p$ denote the polynomial degree of the NURBS curve. The total number of control points $N = m+p+1$ where $m$ is the number of knots  and the degree of NURBS curve. The above definition can be easily extended to higher dimensions.

The B-splines are recursive in polynomial degree, for example, we can derive quadratic B-splines ($p=2$) using linear B-splines ($p=1$), cubic ($p=3$) from quadratic B-splines and so on. Given $m$ knot locations $t_1, t_2, ..., t_{m}$, the B-spline of degree 0 is defined as:
\begin{equation}\label{linearBspline}
B_{i,0}(x) :=
\begin{cases}
1\quad t_i\leq x < t_{i+1},\quad i = 1,2,...,m,\\
0\quad\text{elsewhere.}
\end{cases} 
\end{equation}
Higher order B-splines can then be derived using the recurrence relation \cite{deBoor}:
\begin{equation}\label{NURBS_recurrence}
B_{i,p}(x) := \frac{x-t_i}{t_{i+p} - t_i}B_{i,p-1}(x) + \frac{t_{i+p+1}  -  x}{t_{i+p+1}  -  t_{i+1}}B_{i+1,p-1}(x),\quad p\geq1, i = 1,2,...,m.
\end{equation} 
Something about padding...
These B-splines can be constructed efficiently using the De Boor's algorithm \cite{deBoor}. In Fig. \ref{B-splines}, we show linear, quadratic and cubic splines for $x\in[0,1]$. At the interval boundaries these B-splines go to zero smoothly. The domain of influence of a given control point depends on the respective B-spline (rephrase)??

Next, we describe steps to generate perturbed samples of chord from a given reference chord, $S_{ref}(x)$, using NURBS based parametrization. The first step is to approximate the given reference chord using a NURBS curve with a fixed degree $p$. For this, we need to sample $S_{ref}(x)$ at $N$ locations $\{x_j\}_{j=1}^N$ and compute $B_{i,p}(x_j), \text{ for }i, j = 1,2, ..., N$, such that we can compute the set of control points $\mathbf{c}=\{c_i\}_{i=0}^N$ by solving the following linear system:
\begin{equation}\label{nurbs_inversion}
\mathbf{B}\mathbf{c} = \mathbf{S},
\end{equation}
where $\mathbf{S}\in \mathbb{R}^{N}$ is a vector containing sampled values of the reference curve and $\mathbf{B}\in \mathbb{R}^{N\times N}$ is a matrix with $j$-th row consisting of B-splines values at $x_j$ locations, i.e., $B_{i,p}(x_j), i= 1, 2,..., N$. 
Once the control points are obtained, we can derive the approximate reference curve $S_N(x)$ using \eqref{NURB_curve}. Note that the accuracy of the approximated curve is dependent on the sample locations $x_j$ as well as the degree $p$ and can be set heuristically \cite{adaptiveNurbs}. The number of sampled locations $N$ can be adaptively increased until the following tolerance criteria is met:
\begin{equation}
\frac{||S_N - S_{ref}||}{||S_{ref}||} <\varepsilon.
\end{equation}
More advanced approaches, monotonicity preserving methods, etc??
%\begin{remark}
%How to chose $x$?
%\end{remark}
%The final step is to introduce uncertainty in the above computed control points by associating a probability distribution.
%
%The steps to obtained perturbed samples from a given reference curve is outlined in Fig. \ref{}.
\subsection{Model uncertainty}
\subsection{Wind velocity}

\section{Global sensitivity analysis}\label{sec:GSA}
The objective of sensitivity analysis is to quantify the relative significance of individual inputs (or in combination) and how variations in input values affects the output of interest. In engineering, sensitivity analysis can be employed for a number of reasons: to determine the stability and robustness of a computational model with respect to input parameters, for simplification of stochastic models by fixing the insensitive parameters, and to guide data acquisition campaigns and experimental design to refine the data on sensitive parameters. Sensitivity analysis techniques can be classified as local and global methods, see \cite{RSmith}. In local sensitivity analysis, individual parameters are perturbed around their nominal values allowing for the description of output variability only in a small neighbourhood of nominal input values. Although, local approaches are widely employed due to their ease of implementation and low computational cost, they are unable to quantify global behavior of nonlinearly parametrized models such as aeroelastic models. Global sensitivity approaches, on the other hand, consider the entire range of input values to compute output sensitivities. Therefore, global sensitivity analysis is more suitable for aeroelastic models considered in this work.

\subsection{Sobol analysis}
We employ variance-based Sobol decomposition to perform global sensitivity analysis. This approach allows for quantification of  the relative importance of the input parameters on a scale of $[0,1]$ known as \emph{Sobol indices}.

The main idea of Sobol analysis is to express the total variance of the output in term of contributions from individual parameters and their combinations. For the ease of exposition, let us consider a nonlinear model:
\begin{equation}\label{nonlinear_model}
Y = f(\mathbf{Z}),
\end{equation}
where $\mathbf{Z} = [z_1, z_2, ..., z_M]\in \mathcal{D}_{\mathbf{Z}}\in \mathbb{R}^M$ is the input vector. For simplicity, we assume input parameters are uniformly distributed i.e. $z_i \sim \mathcal{U}(0,1)$ and the support of input set is $\mathcal{D}_\mathbf{Z}  =  [0,1]^M$ where $M$ is the total number input parameters. The Sobol decomposition is defined as:
\begin{equation}\label{sobol_decomp}
f(z_1, z_2, ..., z_M) = f_0+\sum_{i=1}^M f_i(z_i) + \sum_{1\leq<i<j\leq M} f_{ij}(z_i,z_j) + ... + f_{1, 2, ..., M}(z_1, z_2, ..., z_M),
\end{equation}
The above decomposition is only valid for independent input parameters. 
\subsubsection{PCE Ordinary Least Square (PCE\_OLS)}
\subsubsection{PCE Least Angle Regression (PCE\_LAR)}

%\section{Description of test case}

%\section{Sensitivity analysis workflow}

\section{Results}\label{sec:results}

@@

\section{References}
\bibliography{\myreferences}

>>>>

For this case of a rigid turbine, with a uniform inflow field, the main model uncertainties that arise in the BEM formulation that we will discuss in this work are the following \cite{Hansen1993}:
\begin{itemize}
\item The use of (2D) \textit{airfoil polars}; uncertainty arises not only because the actual flow along the blade is 3D, but also because the 2D polars itself can be inaccurate, either when obtained from measurement or from 2D codes like \texttt{XFOIL};
%\item Sheared inflow
\item The assumption of locally 2D flow along the blade is also invalid at the tip; this is typically corrected via \textit{a tip loss model};
\item Three-dimensional flow effects also cause stall delay at the root; this is typically modelled via \textit{a dynamic stall model }and time constant;
\item Yawed inflow is not naturally included in BEM and typically incorporated via \textit{a wake correction}.

@@

\subsection{Geometric uncertainty}
We use Non-Uniform Rational Basis Splines (NURBS) \cite{rogers2000} to perturb the geometrical parameters of the turbine blade, such as the reference chord, twist or experimentally obtained drag-, lift-polar curves. The main advantage of using NURBS is that it provides great flexibility to approximate a large variety of curves with a limited number of control points which can be useful in reducing the number of uncertain parameter. %Further, the set of control points and knots can be directly manipulated to control the smoothness and curvature.
Use of NURBS has already been exploited in many aerodynamics applications, for instance, design optimization of wind-turbine blades, see for e.g. \cite{Ribeiro2012,Bottasso2014} and to parametrize chord and twist curves \cite{Echeverria2017}. In the following, we briefly outline the procedure used to parametrize curves using NURBS.
\subsubsection{NURBS based perturbation}
The value of NURBS curve at each location $x$ is computed using a weighted sum of $n$ basis functions (or B-splines):
\begin{equation}\label{NURB_curve}
S(x)  = \sum_{i=1}^{n} c_i B_{i,p}(x),
\end{equation}
where $S(x)$ is the value of the NURBS curve at location $x$; $c_i$ is the weight of the $i$-th control point and $B_{i,p}(x)$ is the value of B-spline corresponding to the $i$-th control point at $x$ with the subscript $p$ denoting its polynomial degree. A sequence of non-decreasing parameters $X = \{x_1,x_2,...,x_m\}$ called the \textit{knot vector} determines the domain in which any control point is active with the number of elements in the knot vector is given by $m = n+p+1$. B-splines are non-zero only within a knot interval $[x_i,x_{i+1})$ for $i = 1, 2, ..., m-1$ and are zero elsewhere. Therefore, only a few nearby control points are active when computing the NURBS curve \eqref{NURB_curve}. For a curve of polynomial degree p, there are $p+1$ active control points at any location $x$, for e.g., for a linear curve ($p=1$) only two adjacent control points are active at any location. Also, note that the number of B-splines are equal to the number of elements in $X$. Another useful property of the B-splines is that they are recursive in polynomial degree thus we can derive quadratic B-splines using linear B-splines, cubic from quadratic B-splines and so on. Starting with B-spline of degree 0 defined as 
\begin{equation}\label{linearBspline}
B_{i,0}(x) :=
\begin{cases}
1\quad x_i\leq x < x_{i+1},\quad i = 1,2,...,m,\\
0\quad\text{elsewhere,}
\end{cases} 
\end{equation}
we can derive higher order B-splines in the knot interval $[x_{i+p},x_{i+p+1})$ using the recurrence relation \cite{deBoor}:
\begin{equation}\label{NURBS_recurrence}
B_{i,p}(x) := \frac{x - x_i}{x_{i+p} - x_i}B_{i,p-1}(x) + \frac{x_{i+p+1}  -  x}{x_{i+p+1}  -  x_{i+1}}B_{i+1,p-1}(x),\quad p\geq1.
\end{equation} 
These B-splines can be constructed efficiently using the de Boor's algorithm \cite{deBoor}. Note that to perform above computation, we need to pad the knot vector $X$ by repeating the first and last elements $p$ times each. In Fig. \ref{B-splines}, we show linear, quadratic and cubic splines for $x\in[0,1]$ [Also show the padding].

Next, we describe steps to generate perturbed samples of chord from a given reference chord, $S_{ref}(x)$, using NURBS based parametrization. The first step is to approximate the given reference chord using a NURBS curve with a fixed degree $p$ and a limited number of knots $X =\{x_j\}_{j=0}^n$. The knots can be uniformly spaced or chosen heuristically, more advanced approaches for knot selection can be found in, for e.g. \cite{razdan1999knot,LI2005791}. Next we sample $S_{ref}(x)$ at $X$ and also compute $B_{i,p}(x_j), \text{ for }i, j = 1,2, ..., n$.  Now we can formulate a linear system to solve for a set of control points $\mathbf{c}=\{c_i\}_{i=0}^n$ as:
\begin{equation}\label{nurbs_inversion}
\mathbf{B}\mathbf{c} = \mathbf{S}_{ref},
\end{equation}
where $\mathbf{S}_{ref}\in\mathbb{R}^{n}$ is a vector containing sampled values of $S_{ref}(x)$ and $\mathbf{B}\in \mathbb{R}^{n\times n}$ is a matrix with $j$-th row consisting of values of $n$ B-splines sampled at the location $x_j$. Once the control points are obtained, we can derive the approximate reference curve $S_n(x)$ using \eqref{NURB_curve}. In this work, we use uniform random variable to perturb the baseline control points $\mathbf{c}$.

\subsection{Model uncertainty}
\subsection{Wind velocity}
\section{Global sensitivity analysis}\label{sec:GSA}
The objective of sensitivity analysis is to quantify the relative significance of individual inputs (or in combination) and how variations in input values affects the output of interest. In engineering, sensitivity analysis can be employed for a number of reasons: to determine the stability and robustness of a computational model with respect to input parameters, for simplification of stochastic models by fixing the insensitive parameters, and to guide data acquisition campaigns and experimental design to refine the data on sensitive parameters. Sensitivity analysis techniques can be classified as local and global methods, see \cite{RSmith}. In local sensitivity analysis, individual parameters are perturbed around their nominal values allowing for the description of output variability only in a small neighbourhood of nominal input values. Although, local approaches are widely employed due to their ease of implementation and low computational cost, they are unable to quantify global behavior of nonlinearly parametrized models such as aeroelastic models. Global sensitivity approaches, on the other hand, consider the entire range of input values to compute output sensitivities. Therefore, global sensitivity analysis is more suitable for aeroelastic models considered in this work.

\subsection{Sobol analysis}
We employ variance-based Sobol decomposition to perform global sensitivity analysis. This approach allows for quantification of  the relative importance of the input parameters on a scale of $[0,1]$ known as \emph{Sobol indices}. In a Sobol analysis, we express the total variance of the output in term of contributions from individual parameters and their combinations. For the ease of exposition, let us consider a nonlinear model:
\begin{equation}\label{nonlinear_model}
Y = f({z}),
\end{equation}
where ${z} = [z_1, z_2, ..., z_M]\in \mathcal{D}_{z}\in \mathbb{R}^M$ is a vector of random  inputs. For simplicity, we assume input parameters are uniformly distributed i.e. $z_i \sim \mathcal{U}(0,1)$ and the support of input set is $\mathcal{D}_{z}  =  [0,1]^M$ where $M$ is the total number input parameters. The Sobol decomposition is based on the hierarchical representation \cite{RSmith}:
\begin{equation}\label{sobol_decomp}
f(z_1, z_2, ..., z_M) = f_0+\sum_{i=1}^M f_i(z_i) + \sum_{1\leq<i<j\leq M} f_{ij}(z_i,z_j) + ... + f_{1, 2, ..., M}(z_1, z_2, ..., z_M),
\end{equation}
where the zeroth-order function $f_0$ is the mean response of $f$, the first-order univariate functions $f_i(z_i)$ quantify independent contributions due to each random input, second-order functions $f_{i,j}(z_i,z_j)$ represent the effect of interaction between $z_i$ and $z_j$ on the response $f$. Note that the \eqref{sobol_decomp} is only valid for independent input parameters. Higher-order terms are interpreted in a similar manner. Formally, the zeroth-, first- and second-order terms are defined as:
\begin{align}\label{sobol_terms}
f_0 &= \int_{\mathcal{D}_{z}}f(z)dz,\\
f_i(z_i) &= \int_{{\mathcal{D}_{z}}^{M-1}}f(z)dz_{\sim \{i\}} - f_0,\\
f_{ij}(z_i,z_j) &= \int_{{\mathcal{D}_{z}}^{M-2}}f(z)dz_{\sim \{i,j\}} - f_i(z_i)  - f_j(z_j) - f_0,
\end{align}
where ${\mathcal{D}_{z}}^{M-1} = [0,1]^{M-1}$ and the notation $z_{\sim \{i,j\}}$ denotes the vector having all the components of ${z}$ except in the set $\{i,j\}$. We only consider second order expansion of \eqref{sobol_decomp}, as for a number of applications, the second order expansion is usually sufficient and higher-order terms have negligible effect on the output response. In case of wind turbine, suppose $y$ represent the power output that depends on say 3 random inputs:  windspeed $(z_1)$, wind standard deviation $(z_2)$ and RPM $(z_3)$, then $f_0$ represents the mean power output considering all the three random inputs, $f_1(z_1)$ represent the independent contribution of windspeed on the power and $f_{1,2}(z_1,z_2)$ quantify the interactions of wind speed and wind standard deviation on power. Further, the total variance of $f(z)$ is defined as:
\begin{equation}\label{tot_var}
D = \text{Var}[f(z)] = \int_{\mathcal{D}_{z}} f^2(z)dz - f_0^2
\end{equation}
As the expansion in \eqref{sobol_decomp} is not unique, some orthogonality conditions must be imposed, see \cite{Rabitz1999,SOBOL2001271} which allows to express the total variance as
\begin{equation}
D = \sum_{i=1}^M D_i + \sum_{1\leq i<j\leq M} D_{ij},
\end{equation}
where the first- and second-order partial variances are defined as
\begin{align}
D_i &= \int_{0}^{1} f^2_i(z_i)dz_i, \label{partial_variance1}\\
D_{ij} &= \int_{0}^{1} \int_{0}^{1} f^2_{ij}(z_i,z_j)dz_idz_j \label{partial_variance2}. 
\end{align}
The first- and second-order Sobol indices are computed as
\begin{equation}\label{sobol_ind}
S_i = \frac{D_i}{D}, \quad  S_{ij} = \frac{D_{ij}}{D}, \quad i,j=1,2, ..., M,
\end{equation}
and the sum of all indices satisfy
\begin{equation}
\sum_{i=1}^M S_i + \sum_{1\leq i<j\leq M} S_{ij} = 1.
\end{equation}
Finally, the total effect of the parameter $z_i$ on the output response $Y$ is quantified using the total sensitivity indices measure
\begin{equation}
S_{T_i} = S_i + \sum_{j=1}^M S_{ij}.
\end{equation}
The total sensitivity indices can be interpreted as an importance measure for the parameter $z_i$, therefore a large $S_{T_i}$ implies that $z_i$ has a strong influence on $Y$. The computation of $S_{T_i}$ requires the approximation of partial variances defined in \eqref{partial_variance1} - \eqref{partial_variance2}. Using the Monte Carlo method for computing these variance can be prohibitive for computationally expensive models $Y$.  Polynomial Chaos Expansions (PCE) based approaches provides a more efficient alternative for sensitivity analysis. As shown in  \cite{SUDRET2008964}, one can compute the Sobol indices analytically by post-processing the PCE coefficients.
\subsection{PCE based Sobol indices computation}

\subsubsection{Ordinary Least Square (PCE\_OLS)}
\subsubsection{Least Angle Regression (PCE\_LAR)}
%\section{Description of test case}
%\section{Sensitivity analysis workflow}
\section{Results}\label{sec:results}

@@

\newpage
\bibliography{\myreferences}
